---
title: "Reproducible Research PA1"
author: "J. Voltz"
date: "February 20, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Loading and preprocessing the data


```{r, echo=TRUE}
activity <- read.csv("activity.csv", header = T)

# Review
dim(activity)
```



```{r, echo=TRUE}
str(activity)
```

Date class and interval factor
```{r, echo=TRUE}
activity$date <- as.Date(activity$date, format = "%Y-%m-%d")
activity$interval <- factor(activity$interval)

##missing data
missing_activity<- activity[is.na(activity$steps),]
dim(missing_activity)
```
What is the mean total number of steps taken per day?
  
  Calculate the total number of steps taken per day
```{r, echo=TRUE}
activity1 <- activity[!is.na(activity$steps),]

TotalSteps <- with(activity, tapply(steps, as.factor(activity$date), sum, na.rm = TRUE))

hist(TotalSteps, main = "Total Steps Taken per Day", xlab = "Total number of steps")
```
Calculate and report the mean and median of the total number of steps taken per day
```{r, echo=TRUE}
summary(TotalSteps)
```
What is the average daily activity pattern?
  
  Make a time series plot (i.e. type = "l") of the 5-minute   interval (x-axis) and the average number of steps taken,   averaged across all days (y-axis)
```{r, echo=TRUE}
mean_steps <- with(activity1, tapply(steps, activity1$interval, mean))
interval<- levels(as.factor(activity$interval))
```

```{r, echo=TRUE}
plot(interval, mean_steps, type = "l", xlab = "Time Interval (5 mins.)", ylab = "Mean Steps Taken (all Days)", main = "Average Steps Taken at 5 minute Intervals",  col = "black")
```
Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r, echo=TRUE}
table <- data.frame(mean_steps, interval)
table[table$mean_steps==max(table$mean_steps),][2]
```
Imputing missing values

  Calculate and report the total number of missing values    in the dataset (i.e. the total number of rows with NAs)
```{r, echo=TRUE}
length(missing_activity$steps)
```
Devise a strategy for filling in all of the missing values in the dataset
```{r, echo=TRUE}
mean_steps <- with(activity1, tapply(steps, activity1$interval, mean))
missing_activity$steps <- mean_steps
```
Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r, echo=TRUE}
new_activity <- rbind(activity1, missing_activity)
new_activity <- new_activity[order(new_activity$date), ]
```
Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day
```{r, echo=TRUE}
TotalSteps2 <- with(new_activity, tapply(steps, as.factor(new_activity$date), sum))

hist(TotalSteps2, main = "Total Steps Taken per Day", xlab = "Total number of steps")
```

    Calculate and report the mean and median of the total number of steps taken per day.

Mean and median total number of steps taken per day WITHOUT filling in the missing values
```{r, echo=TRUE}
##With NA
summary(TotalSteps)
##Without NA
summary(TotalSteps2)
```
Are there differences in activity patterns between weekdays and weekends?
```{r, echo=TRUE}
activity$day <- as.factor(weekdays(activity$date))

activity$is_weekday <- ifelse(!(activity$day %in% c("Saturday","Sunday")), TRUE, FALSE) 

weekdays <- activity[activity$is_weekday,]
interval_weekdays <- aggregate(weekdays$steps, by=list(interval=weekdays$interval), FUN=mean)

weekends <- activity[!activity$is_weekday,]
interval_weekends <- aggregate(weekends$steps, by=list(interval=weekends$interval), FUN=mean)

colnames(interval_weekdays) <- c("interval", "avgsteps")
colnames(interval_weekends) <- c("interval", "avgsteps")

interval_weekdays$day <- "Weekday"
interval_weekends$day <- "Weekend"

week <- rbind(interval_weekends, interval_weekdays)

week$day <- as.factor(week$day)
```

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)
```{r, echo=TRUE}
library(lattice)
xyplot(mean_steps ~  interval | day, data = week, layout = c(1,2), type ="l", ylab="Steps")
```


